---
title: "Single strain deterministic and stochastic models"
author: "Anastasia Chatzilena"
---

```{r setup, include=FALSE}
library(deSolve)
library(dplyr)
library(rstan)
library(outbreaks)
library(bayesplot)

rstan_options(auto_write = TRUE)           
options(mc.cores = parallel::detectCores())

knitr::opts_chunk$set(cache = TRUE, echo = TRUE, message = FALSE, warning = FALSE)
```

In this example, we examine an outbreak of influenza A (H1N1) in 1978 at a British boarding school. We use daily data on the number of students in bed, spanning over a time interval of 14 days. There were 763 male students which were mostly full boarders and 512 of them became ill. The outbreak lasted from the 22nd of January to the 4th of February and it is reported that one infected boy started the epidemic and then it spread rapidly in the relatively closed community of the boarding school. The data are freely available in the R package outbreaks, maintained as part of the R Epidemics Consortium (RECON; \url{http://www.repidemicsconsortium.org}).

Load data of influenza:

```{r load_flu_data, echo = TRUE}
onset <- influenza_england_1978_school$date    
cases <- influenza_england_1978_school$in_bed  #Number of students in bed
N = length(onset) # Number of days observed throughout the outbreak
pop = 763         # Population 
sample_time=1:N

# Modify data into a form suitable for Stan
flu_data = list(n_obs = N,
                n_theta = 2,
                n_difeq = 3,
                n_pop = pop,
                y = cases,
                t0 = 0,
                ts = sample_time)

# Specify parameters to monitor

parameters = c("y_hat", "y_init", "theta",  "R_0")  #deterministic models (Model 1, Model 2)

parameters_stoch = c("y_hat", "y_init", "theta", "kappa", "lambda", "phi", "s_sq", "sigma", "R_0")  #stochastic model (Model 3)
```

In the context of a deterministic SIR, in order to estimate the transmission and the recovery rate, we use two different specifications and we fit the models using Stan's NUTS algorithm and the mean-field ADVI variant.

## Model 1 - Poisson model

Stan model statement:

```{r stan_model1, echo=TRUE}
mod1_stat <- '
functions {
  real[] SIR(real t,  // time
  real[] y,           // system state {susceptible,infected,recovered}
  real[] theta,       // parameters
  real[] x_r,
  int[] x_i) {
  
  real dy_dt[3];
  
  dy_dt[1] = - theta[1] * y[1] * y[2];
  dy_dt[2] = theta[1] * y[1] * y[2] - theta[2] * y[2];
  dy_dt[3] = theta[2] * y[2];
  
  return dy_dt;
  }
  
  }
  data {
  int<lower = 1> n_obs;       // number of days observed
  int<lower = 1> n_theta;     // number of model parameters
  int<lower = 1> n_difeq;     // number of differential equations
  int<lower = 1> n_pop;       // population 
  int y[n_obs];           // data, total number of infected individuals each day
  real t0;                // initial time point (zero)
  real ts[n_obs];         // time points observed
  }
  
  transformed data {
  real x_r[0];
  int x_i[0];
  }
  
  parameters {
  real<lower = 0> theta[n_theta]; // model parameters {beta,gamma}
  real<lower = 0, upper = 1> S0;  // initial fraction of susceptible individuals
  }
  
  transformed parameters{
  real y_hat[n_obs, n_difeq]; // solution from the ODE solver
  real y_init[n_difeq];     // initial conditions for both fractions of S and I
  
  y_init[1] = S0;
  y_init[2] = 1 - S0;
  y_init[3] = 0;
  y_hat = integrate_ode_rk45(SIR, y_init, t0, ts, theta, x_r, x_i);
  
  }
  
  model {
  real lambda[n_obs];      //poisson parameter
  
  //priors
  theta[1] ~ lognormal(0,1);
  theta[2] ~ gamma(0.004,0.02);  //Assume mean infectious period = 5 days 
  S0 ~ beta(0.5, 0.5);
  
  //likelihood
  for (i in 1:n_obs){
  lambda[i] = y_hat[i,2]*n_pop;
  }
  y ~ poisson(lambda);
  }
  
  generated quantities {
  real R_0;      // Basic reproduction number
  R_0 = theta[1]/theta[2];
  }
'

m1 <- stan_model(model_code = mod1_stat)
```

Fit and sample from the posterior using Hamiltonian Monte Carlo-NUTS:

```{r NUTS_fit1, dependson=c(load_flu_data) , echo=FALSE}
n_chains=5
n_warmups=500
n_iter=100500
n_thin=50
set.seed(1234)
# Set initial values:
ini_1 = function(){
  list(theta=c(runif(1,0,5), runif(1,0.2,0.4)), 
       S0=runif(1,(pop-3)/pop,(pop-1)/pop))  
}

time.start_nuts1 <- Sys.time()
nuts_fit_1 = sampling(m1, data = flu_data, pars = parameters, init = ini_1, chains = n_chains, warmup = n_warmups, iter = n_iter, thin=n_thin, seed=13219)
time.end_nuts1 <- Sys.time()
duration_nuts1<- time.end_nuts1 - time.start_nuts1

nuts_fit_1_summary <- summary(nuts_fit_1, pars = c("lp__", "theta[1]", "theta[2]", "y_init[1]", "R_0"))$summary
print(nuts_fit_1_summary,scientific=FALSE,digits=2)
posts_1 <-  rstan::extract(nuts_fit_1)
```

Check HMC diagnostics:

```{r diagnostics1, echo=FALSE}
mod1_diagnostics <-rstan::get_sampler_params(nuts_fit_1)

# Check for divergent transitions
rstan::check_divergences(nuts_fit_1)

posterior_1 <- as.array(nuts_fit_1)
color_scheme_set("viridis")
# Markov chain traceplots
mcmc_trace(posterior_1, pars="lp__")
mcmc_trace(posterior_1, pars=c("theta[1]", "theta[2]", "y_init[1]"))
mcmc_trace(posterior_1, pars="R_0")

# Univariate and bivariate marginal posterior distributions
pairs(nuts_fit_1, pars = c("theta[1]", "theta[2]", "y_init[1]"), labels = c("beta", "gamma", "s(0)"), 
      cex.labels=1.5, font.labels=9, condition = "accept_stat__")  

# Kernel density estimates of each Markov chain separately, overlaid
mcmc_dens_overlay(posterior_1, pars=c("theta[1]", "theta[2]", "y_init[1]"))

#Central posterior uncertainty intervals
mcmc_intervals(posterior_1,pars = c("theta[1]", "theta[2]", "y_init[1]"))
```

Plot model fit, median and 95% credible interval:

```{r plot_nuts_fit1, echo=FALSE}
# Model fitted values across the observed time period
fit_I_1 <- posts_1$y_hat[,,2]    # Fitted fraction of infected 
fit_SIR_1 <- fit_I_1*pop         # Fitted number of infected
median_I_1 = apply(fit_SIR_1, 2, median)
low_I_1 = apply(fit_SIR_1, 2, quantile, probs=c(0.025))
high_I_1 = apply(fit_SIR_1, 2, quantile, probs=c(0.975))
df_sample_N = data.frame(cases, sample_time)
df_fit_I_1 = data.frame(median_I_1, low_I_1, high_I_1, sample_time)

save(df_sample_N,file="data.Rda")
save(df_fit_I_1,file="df_I_det_Poiss.Rda")

ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_I_1, ymax = high_I_1), fill = "orange", alpha = 0.6) +
  geom_line(data = df_fit_I_1, aes(x=sample_time, y=median_I_1, color = "Median"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Median'='darkorange3'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA),  linetype=c(0,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw()+ theme(text = element_text(size=20))
```

Approximate the posterior using ADVI(mean-field):

```{r ADVI_1, dependson=c(load_flu_data), echo=FALSE}
set.seed(1234)
# Set initial values:
ini_vb_1 = function(){
  list(params=c(runif(1,1.85,1.92), runif(1,0.47,0.49)), 
       S0=runif(1,(pop-2)/pop,(pop-1)/pop))}

time.start_vb_1 <- Sys.time()
vb_fit_1=vb(m1, data = flu_data, pars = parameters, init = ini_vb_1, iter = 10000, tol_rel_obj = 0.001, seed=16735679)
time.end_vb_1 <- Sys.time()
vb_duration_1 <- time.end_vb_1 - time.start_vb_1

vb_fit_1_summary <- summary(vb_fit_1, pars = c("lp__", "theta[1]", "theta[2]", "y_init[1]", "R_0"))$summary
print(vb_fit_1_summary,scientific=FALSE,digits=2)
posts_vb_1 <-  rstan::extract(vb_fit_1)
```

Plot model fit, median and 95% credible interval:

```{r plot_advi_1, echo=FALSE}
# Model fitted values across the observed time period
fit_I_vb_1 <- posts_vb_1$y_hat[,,2]
fit_SIR_vb_1 <- fit_I_vb_1*pop
median_I_vb_1 = apply(fit_SIR_vb_1, 2, median)
low_I_vb_1 = apply(fit_SIR_vb_1, 2, quantile, probs=c(0.025))
high_I_vb_1 = apply(fit_SIR_vb_1, 2, quantile, probs=c(0.975))
df_fit_I_vb_1 = data.frame(median_I_vb_1, low_I_vb_1, high_I_vb_1, sample_time)
save(df_fit_I_vb_1,file="df_I_det_Poiss_vb.Rda")

ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_I_vb_1, ymax = high_I_vb_1), fill = "light green", alpha = 0.6) +
  geom_line(data = df_fit_I_vb_1, aes(x=sample_time, y=median_I_vb_1, color = "Median"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Median'='dark green'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA), linetype=c(0,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw()+theme(text = element_text(size=20))
```

## Model 2 - Binomial model

Stan model statement:

```{r stan_model2, echo=TRUE}
mod2_stat <- '
functions {
  real[] SIR(real t,  // time
  real[] y,           // system state {susceptible,infected,recovered}
  real[] theta,       // parameters
  real[] x_r,
  int[] x_i) {
  
  real dy_dt[3];
  
  dy_dt[1] = - theta[1] * y[1] * y[2];
  dy_dt[2] = theta[1] * y[1] * y[2] - theta[2] * y[2];
  dy_dt[3] = theta[2] * y[2];
  
  return dy_dt;
  }
  
  }
  data {
  int<lower = 1> n_obs;       // number of days observed
  int<lower = 1> n_theta;     // number of model parameters
  int<lower = 1> n_difeq;     // number of differential equations
  int<lower = 1> n_pop;       // population 
  int y[n_obs];           // data, total number of infected individuals each day
  real t0;                // initial time point (zero)
  real ts[n_obs];         // time points observed
  }
  
  transformed data {
  real x_r[0];
  int x_i[0];
  }
  
  parameters {
  real<lower = 0> theta[n_theta]; // model parameters {beta,gamma}
  real<lower = 0, upper = 1> S0;  // initial fraction of susceptible individuals
  }
  
  transformed parameters{
  real y_hat[n_obs, n_difeq]; // solution from the ODE solver
  real y_init[n_difeq];     // initial conditions for both fractions of S and I
  
  y_init[1] = S0;
  y_init[2] = 1 - S0;
  y_init[3] = 0;
  y_hat = integrate_ode_rk45(SIR, y_init, t0, ts, theta, x_r, x_i);
  
  }
  
  model {

  //priors
  theta[1] ~ lognormal(0,1);
  theta[2] ~ gamma(0.004,0.02);  //Assume mean infectious period = 5 days 
  S0 ~ beta(0.5, 0.5);
  
  //likelihood
  y ~ binomial(n_pop, y_hat[, 2]);
  }
  
  generated quantities {
  real R_0;      // Basic reproduction number
  R_0 = theta[1]/theta[2];
  }
'

m2 <- stan_model(model_code = mod2_stat)
```

Fit and sample from the posterior using Hamiltonian Monte Carlo-NUTS:

```{r NUTS_fit2, dependson=c(load_flu_data) , echo=FALSE}
n_chains=5
n_warmups=500
n_iter=100500
n_thin=50
set.seed(1234)
# Set initial values:
ini_2 = function(){
  list(theta=c(runif(1,0,5), runif(1,0.2,0.4)), 
       S0=runif(1,(pop-3)/pop,(pop-1)/pop))  
}

time.start_nuts2 <- Sys.time()
nuts_fit_2 = sampling(m2, data = flu_data, pars = parameters, init = ini_2, chains = n_chains, warmup = n_warmups, iter = n_iter, thin=n_thin, seed=259158591)
time.end_nuts2 <- Sys.time()
duration_nuts2 <- time.end_nuts2 - time.start_nuts2

nuts_fit_2_summary <- summary(nuts_fit_2, pars = c("lp__", "theta[1]", "theta[2]", "y_init[1]", "R_0"))$summary
print(nuts_fit_2_summary,scientific=FALSE,digits=2)
posts_2 <-  rstan::extract(nuts_fit_2)
```

Check HMC diagnostics:

```{r diagnostics2, echo=FALSE}
mod2_diagnostics <-rstan::get_sampler_params(nuts_fit_2)

# Check for divergent transitions
rstan::check_divergences(nuts_fit_2)

posterior_2 <- as.array(nuts_fit_2)
color_scheme_set("viridis")
# Markov chain traceplots
mcmc_trace(posterior_2, pars="lp__")
mcmc_trace(posterior_2, pars=c("theta[1]", "theta[2]", "y_init[1]"))
mcmc_trace(posterior_2, pars="R_0")

# Univariate and bivariate marginal posterior distributions
pairs(nuts_fit_2, pars = c("theta[1]", "theta[2]", "y_init[1]"), labels = c("beta", "gamma", "s(0)"), 
      cex.labels=1.5, font.labels=9, condition = "accept_stat__")  

# Kernel density estimates of each Markov chain separately, overlaid
mcmc_dens_overlay(posterior_2, pars=c("theta[1]", "theta[2]", "y_init[1]"))

#Central posterior uncertainty intervals
mcmc_intervals(posterior_2,pars = c("theta[1]", "theta[2]", "y_init[1]"))
```

Plot model fit, median and 95% credible interval:

```{r plot_nuts_fit2, echo=FALSE}
# Model fitted values across the observed time period
fit_I_2 <- posts_2$y_hat[,,2]    # Fitted fraction of infected 
fit_SIR_2 <- fit_I_2*pop         # Fitted number of infected
median_I_2 = apply(fit_SIR_2, 2, median)
low_I_2 = apply(fit_SIR_2, 2, quantile, probs=c(0.025))
high_I_2 = apply(fit_SIR_2, 2, quantile, probs=c(0.975))
df_fit_I_2 = data.frame(median_I_2, low_I_2, high_I_2, sample_time)

ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_I_2, ymax = high_I_2), fill = "orange", alpha = 0.6) +
  geom_line(data = df_fit_I_2, aes(x=sample_time, y=median_I_2, color = "Median"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Median'='darkorange3'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA),  linetype=c(0,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw()+ theme(text = element_text(size=20))
```

Approximate the posterior using ADVI(mean-field):

```{r ADVI_2, dependson=c(load_flu_data), echo=FALSE}
set.seed(1234)
# Set initial values:
ini_vb_2 = function(){
  list(params=c(runif(1,1.85,1.92), runif(1,0.47,0.49)),
       S0=runif(1,(pop-2)/pop,(pop-1)/pop))}

time.start_vb_2 <- Sys.time()
vb_fit_2=vb(m2, data = flu_data, pars = parameters, init = ini_vb_2, iter = 10000, tol_rel_obj = 0.001, seed=121671623)
time.end_vb_2 <- Sys.time()
vb_duration_2 <- time.end_vb_2 - time.start_vb_2

vb_fit_2_summary <- summary(vb_fit_2, pars = c("lp__", "theta[1]", "theta[2]", "y_init[1]", "R_0"))$summary
print(vb_fit_2_summary,scientific=FALSE,digits=2)
posts_vb_2 <-  rstan::extract(vb_fit_2)
```

Plot model fit, median and 95% credible interval:

```{r plot_advi2, echo=FALSE}
# Model fitted values across the observed time period
fit_I_vb_2 <- posts_vb_2$y_hat[,,2]
fit_SIR_vb_2 <- fit_I_vb_2*pop
median_I_vb_2 = apply(fit_SIR_vb_2, 2, median)
low_I_vb_2 = apply(fit_SIR_vb_2, 2, quantile, probs=c(0.025))
high_I_vb_2 = apply(fit_SIR_vb_2, 2, quantile, probs=c(0.975))
df_fit_I_vb_2 = data.frame(median_I_vb_2, low_I_vb_2, high_I_vb_2, sample_time)

ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_I_vb_2, ymax = high_I_vb_2), fill = "light green", alpha = 0.6) +
  geom_line(data = df_fit_I_vb_2, aes(x=sample_time, y=median_I_vb_2, color = "Median"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Median'='dark green'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA), linetype=c(0,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw()+theme(text = element_text(size=20))
```

Estimates of the Poisson model (Model 1) are very close to the estimates of the Binomial model (Model 2), irrespectively of the inference method used. In both models, ADVI results in narrower credible intervals for $\beta$ and $R_0$ compared to NUTS, suggesting that ADVI may be underestimating the posterior uncertainty, as has been observed in the past.
Both deterministic models have a reasonable fit to the data but underestimate the overall uncertainty thus resulting in overly precise estimates which fail to capture the data appropriately.

## Model 3 - Poisson stochastic model

Even though the deterministic approach gives us an insight into the dynamics of the disease, adding stochasticity may allow for a more accurate estimation of the parameters related to the spread of the disease, as the stochastic component can absorb the noise generated by a possible mis-specification of the model.
We extend our Poisson model to include stochastic variation by incorporating an Ornstein-Uhlenbeck (OU) process.

Stan model statement:

```{r stan_model3, echo=TRUE}
mod3_stat <- '
functions {

  real[] SIR(real t,  // time
  real[] y,           // system state {susceptible,infected,recovered}
  real[] theta,       // parameters 
  real[] x_r,
  int[] x_i) {
  
  real dy_dt[3];
  
  dy_dt[1] = - theta[1] * y[1] * y[2];
  dy_dt[2] = theta[1] * y[1] * y[2] - theta[2] * y[2];
  dy_dt[3] = theta[2] * y[2];
  
  return dy_dt;
  }
  
  }
  
  data {
  int<lower = 1> n_obs;       // number of days observed
  int<lower = 1> n_theta;     // number of model parameters
  int<lower = 1> n_difeq;     // number of differential equations
  int<lower = 1> n_pop;       // population 
  int y[n_obs];           // data, total number of infected individuals each day
  real t0;                // initial time point (zero)
  real ts[n_obs];         // time points that were observed
  }
  
  transformed data {
  real x_r[0];
  int x_i[0];
  }
  
  parameters {
  real<lower = 0> theta[n_theta]; // model parameters {beta,gamma}
  real<lower = 0, upper = 1> S0;  // initial fraction of susceptible individuals
  real kappa[n_obs];              // logarithm of poisson parameter
  real<lower=0> phi;              // speed of reversion
  real<lower=0> s_sq;             // square of instantaneous diffusion term
  }
  
  transformed parameters{
  real y_hat[n_obs, n_difeq]; // solution from the ODE solver
  real y_init[n_difeq];       // initial conditions for both fractions of S and I
  real sigma;                  // variance of OU process
  real lambda[n_obs];          // poisson parameter
  
  y_init[1] = S0;
  y_init[2] = 1 - S0;
  y_init[3] = 0;
  
  y_hat = integrate_ode_rk45(SIR, y_init, t0, ts, theta, x_r, x_i);
  
  for (i in 1:n_obs){
  lambda[i]=exp(kappa[i]);
  }
  
  sigma=(1-exp(-2*phi))*(s_sq/(2*phi));
  }
  
  model {
  
  real mu[n_obs];    // logarithm of solution of deterministic model
  
  //priors
  theta[1] ~ lognormal(0,1); 
  theta[2] ~ gamma(0.004,0.02); //Assume mean infectious period = 5 days
  S0 ~ beta(0.5, 0.5);
  phi ~ normal(0,10);
  s_sq ~ inv_gamma(0.1,0.1);
  
  //likelihood
  for (i in 1:n_obs){
  mu[i] = log(y_hat[i,2]*n_pop);
  }
  
  kappa[1]~normal(mu[1],sigma); // kappa=log(lambda) is an Ornstein-Uhlenbeck process
  for (i in 2:n_obs){
  kappa[i]~normal(mu[i]+(kappa[i-1]-mu[i])*exp(-phi),sigma);   
  }
  
  y ~ poisson(lambda);
  }
  
  generated quantities {
  real R_0;      // Basic reproduction number
  R_0 = theta[1]/theta[2];
  }
'

m3 <- stan_model(model_code = mod3_stat)
```

Fit and sample from the posterior using Hamiltonian Monte Carlo-NUTS:

```{r NUTS_fit3, dependson=c(load_flu_data) , echo=FALSE}
n_chains=5
n_warmups=500
n_iter=100500
n_thin=50
set.seed(123)
# Set initial values
ini_3 = function(){
  list(params=c(runif(1,1,4), runif(1,0.2,0.6)), S0=runif(1,(pop-3)/pop,(pop-1)/pop), 
       phi=runif(1,0.5,6.5), s_sq=runif(1,0.5,4.5))
}

# Test / debug the model
test = sampling(m3, data = flu_data, pars = parameters_stoch, init = ini_3, chains = 1, iter = 10)

time.start_nuts3 <- Sys.time()
nuts_fit_3 = sampling(m3, data = flu_data, pars = parameters_stoch, init = ini_3, chains = n_chains, warmup = n_warmups, iter = n_iter, thin=n_thin, seed=13219)
time.end_nuts3 <- Sys.time()
duration_nuts3 <- time.end_nuts3 - time.start_nuts3

nuts_fit_3_summary <- summary(nuts_fit_3, pars = c("lp__", "theta[1]", "theta[2]", "y_init[1]", "R_0", "phi", "s_sq"))$summary
print(nuts_fit_3_summary,scientific=FALSE,digits=2)
print(duration_nuts3)
posts_3 <-  rstan::extract(nuts_fit_3)
```

Check HMC diagnostics:

```{r diagnostics3, echo=FALSE}
mod3_diagnostics <-rstan::get_sampler_params(nuts_fit_3)

# Check for divergent transitions
rstan::check_divergences(nuts_fit_3)

posterior_3 <- as.array(nuts_fit_3)
# Markov chain traceplots
mcmc_trace(posterior_3, pars="lp__")
mcmc_trace(posterior_3, pars=c("theta[1]", "theta[2]", "y_init[1]"))
mcmc_trace(posterior_3, pars=c("phi","s_sq"))
mcmc_trace(posterior_3, pars="R_0")

pairs(nuts_fit_3, pars = c("theta[1]", "theta[2]", "y_init[1]", "phi","s_sq"), labels = c("beta", "gamma", "s(0)", "phi", "s_sq"),
      cex.labels=1.5, font.labels=9, condition = "accept_stat__")

mcmc_dens_overlay(posterior_3, pars=c("theta[1]", "theta[2]", "y_init[1]", "phi","s_sq"))
mcmc_intervals(posterior_3,pars = c("theta[1]", "theta[2]", "y_init[1]", "phi","s_sq")) 
```

Plot deterministic and stochastic model fit, medians and 95% credible intervals:

```{r plot_nuts_fit3, echo=FALSE}
# Model fitted values
fit_lambda <- posts_3$lambda
post_kappa <-posts_3$kappa
fit_kappa <- unlist(post_kappa)
fit_I_3 <- posts_3$y_hat[,,2]
fit_SIR_3 <- fit_I_3*pop
post_mu <- log(fit_I_3*pop)
fit_mu <- unlist(post_mu)
fit_phi <- posts_3$phi
fit_sigma <- posts_3$sigma
mean_lambda <- matrix(nrow=10000, ncol=14)   # mean of Poisson parameter
mean_lambda[,1]=exp(fit_mu[,1]+(fit_sigma/2));
for (i in 2:N){
  mean_lambda[,i]=exp(fit_mu[,i]+(fit_kappa[,i-1]-fit_mu[,i])*exp(-fit_phi) +(fit_sigma/2));
}
median_lmean = apply(mean_lambda, 2, median)
low_lmean = apply(mean_lambda, 2, quantile, probs=c(0.025))
high_lmean = apply(mean_lambda, 2, quantile, probs=c(0.975))
df_fit_lmean = data.frame(median_lmean, low_lmean, high_lmean, sample_time)

load("df_I_det_Poiss.Rda") # fitted values from the deterministic model
median_I<-df_fit_I_1$median_I_1
low_I<-df_fit_I_1$low_I_1
high_I<-df_fit_I_1$high_I_1
df_fit_I = data.frame(median_I, low_I, high_I, sample_time)

# Plot of the data along with the fitted mean of Poisson parameter and the fitted number of infected from the deterministic model 
ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_lmean, ymax = high_lmean), fill = "lightblue", alpha = 0.6) +
  geom_ribbon(aes(x=sample_time, ymin = low_I, ymax = high_I), fill = "orange", alpha = 0.6) +
  geom_line(data = df_fit_lmean, aes(x=sample_time, y=median_lmean, color="Fitted stochastic model"),size = 1.3) +
  geom_line(data = df_fit_I, aes(x=sample_time, y=median_I, color = "Fitted deterministic model"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Fitted stochastic model'='lightblue4', 'Fitted deterministic model'='darkorange3'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA,NA), linetype=c(0,1,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw()+theme(text = element_text(size=20))
```

Approximate the posterior using ADVI(mean-field):

```{r ADVI_3, dependson=c(load_flu_data), echo=FALSE}
set.seed(123)
# Set initial values:
ini_vb_3 = function(){
  list(params=c(runif(1,1.9,2.1), runif(1,0.5,0.56)),
       S0=runif(1,(pop-3)/pop,(pop-1)/pop),
       phi=runif(1,0.9,6.1), s_sq=runif(1,0.59,3.4))
}

time.start_vb_3 <- Sys.time()
vb_fit_3=vb(m3, data = flu_data, pars = parameters_stoch, init = ini_vb_3, iter = 10000, tol_rel_obj = 0.001, seed=24176217)
time.end_vb_3 <- Sys.time()
vb_duration_3 <- time.end_vb_3 - time.start_vb_3

vb_fit_3_summary <- summary(vb_fit_3, pars = c("theta[1]", "theta[2]", "y_init[1]", "R_0", "phi", "s_sq"))$summary
print(vb_fit_3_summary,scientific=TRUE,digits=2)
print(vb_duration_3)
posts_vb_3 <-  rstan::extract(vb_fit_3)
```

Plot deterministic and stochastic model fit, medians and 95% credible intervals:

```{r plot_advi3, echo=FALSE}
# Model fitted values across the observed time period
fit_lambda_vb <- posts_vb_3$lambda
post_kappa_vb <- posts_vb_3$kappa
fit_kappa_vb <- unlist(post_kappa_vb)
fit_I_vb_3 <- posts_vb_3$y_hat[,,2]
fit_SIR_vb_3 <- fit_I_vb_3*pop
post_mu_vb <- log(fit_I_vb_3*pop)
fit_mu_vb <- unlist(post_mu_vb)
fit_phi_vb <- posts_vb_3$phi
fit_sigma_vb <- posts_vb_3$sigma
mean_lambda_vb <- matrix(nrow=10000, ncol=14)
mean_lambda_vb[,1]=exp(fit_mu_vb[,1]+(fit_sigma_vb/2));  # mean of Poisson parameter
for (i in 2:N){
  mean_lambda_vb[,i]=exp(fit_mu_vb[,i]+(fit_kappa_vb[,i-1]- fit_mu_vb[,i])*exp(-fit_phi_vb)+(fit_sigma_vb/2));
}
median_lmean_vb = apply(mean_lambda_vb, 2, median)
low_lmean_vb = apply(mean_lambda_vb, 2, quantile, probs=c(0.025))
high_lmean_vb = apply(mean_lambda_vb, 2, quantile, probs=c(0.975))
df_fit_lmean_vb = data.frame(median_lmean_vb, low_lmean_vb, high_lmean_vb, sample_time)
load("df_I_det_Poiss_vb.Rda")  # fitted values from the deterministic model

median_I_vb<-df_fit_I_vb_1$median_I_vb_1
low_I_vb<-df_fit_I_vb_1$low_I_vb_1
high_I_vb<-df_fit_I_vb_1$high_I_vb_1
df_fit_I_vb <-data_frame(median_I_vb, low_I_vb, high_I_vb)

# Plot of the data along with the fitted mean of Poisson parameter and the fitted number of infected from the deterministic model
ggplot(df_sample_N, aes(x=sample_time, y=cases)) +
  geom_ribbon(aes(x=sample_time, ymin = low_lmean_vb, ymax = high_lmean_vb), fill = "lightblue", alpha = 0.6) +
  geom_ribbon(aes(x=sample_time, ymin = low_I_vb, ymax = high_I_vb), fill = "orange", alpha = 0.6) +
  geom_line(data = df_fit_lmean_vb, aes(x=sample_time, y=median_lmean_vb, color="Fitted stochastic model"),size = 1.3) +
  geom_line(data = df_fit_I_vb, aes(x=sample_time, y=median_I_vb, color = "Fitted deterministic model"), size = 1.3) +
  geom_point(shape = 19, size = 3, (aes(color="Data"))) +
  scale_colour_manual(name='', values=c('Data'='black', 'Fitted stochastic model'='lightblue4', 'Fitted deterministic model'='darkorange3'))+
  guides(colour = guide_legend(override.aes = list(shape=c(16,NA,NA),linetype=c(0,1,1))))+
  labs(x = "Time (days)", y = "Number of Infected students") + 
  scale_x_continuous(limits=c(0, 14), breaks=c(0,7,14)) +
  scale_y_continuous(limits=c(0,400), breaks=c(0,100,200,300,400)) +
  theme_bw() + theme(text = element_text(size=20))
```

The resulting 95$\%$ credible intervals from ADVI have shorter length compared to NUTS.
Also, adding stochasticity improves the fit to the data.


Summing up, the results of both the deterministic and the stochastic setting bring us to the preliminary conclusion that if we are interested in real-time inference both methods are feasible and efficient. In terms of computational time ADVI is extremely efficient.